---
title: "Microbiome Fruit - Beta Diversity"
author: "Shyan Polman"
date: "`r Sys.Date()`" 
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)

```
# 3. Beta Diversity

## 3.1 CLR transformation
### 3.1.1 Replace zeros with pseudocount 0 = 0.001
```{r}
asv_zerorep <- asv_16S_all
asv_zerorep[asv_zerorep == 0] <- 0.001
# tail(asv_zerorep)

species_zerorep <- species_all
species_zerorep[species_zerorep == 0] <- 0.001

genus_zerorep <- genus_all
genus_zerorep[genus_zerorep == 0] <- 0.001

filtered_asv_20_zerorep <- filtered_asv_20
filtered_asv_20_zerorep[filtered_asv_20_zerorep == 0] <- 0.001


```
### 3.1.2 Create function with clr formula to apply transformation (for clr formula, samples = columns)
```{r}
clr <- function(x) sweep(log(x), 1, rowMeans(log(x)), "-")

# log(x): calculates the natural logarithm of each element in the matrix x; log() computes the natural logarithm
# rowMeans(log(x)): computes the mean of each row in the matrix log(x). rowMeans() calculates the row-wise means
# sweep(log(x), 1, rowMeans(log(x)), "-"): This is where the sweep() function is used - subtracts the row-wise means (obtained from step 2) from each element in the matrix log(x) along the first dimension (rows) - the 1 parameter specifies the dimension along which the operation should be performed, and "-" indicates subtraction

```
### 3.1.3 Apply clr-transformation function to zero-replaced tables
```{r}
asv_tx <- data.frame(t(clr(t(asv_zerorep))), check.names = F)

species_tx <- data.frame(t(clr(t(species_zerorep))), check.names = F)

genus_tx <- data.frame(t(clr(t(genus_zerorep))), check.names = F)

transformed_filtered_asv_20 <- data.frame(t(clr(t(filtered_asv_20_zerorep))), check.names = F)

```
### 3.1.4 Subset feature data into Treatment or Day groups if needed
Assign 'group' with desired subgroup variable (see microbiome-ocp.Rmd)
```{r}

group <- map_all[,1:4]
head(group)

group$treatment <- factor(group$treatment)
group$week <- factor(group$week)

```

```{r}
# subset clr-transformed data according to 'group', if needed 
asv_tx <- asv_tx[, colnames(asv_tx) %in% group$SampleID]

```
### 3.1.5 NMDS
transpose data frames (samples = rows) and make matrix
```{r}
clr_m <- as.matrix(t(asv_tx))

```
check how many dimensions to use for NMDS
```{r}
clr_nmds_2 = metaMDS(clr_m, distance = "euclidean", parallel = 2, k = 2)
clr_nmds_3 = metaMDS(clr_m, distance = "euclidean", parallel = 2, k = 3)
clr_nmds_2$stress

```
extract NMDS scores
```{r}
clr_scores = as.data.frame(scores(clr_nmds_3))

```
add map data to scores
```{r}
clr_scores$treatment <- group$treatment
clr_scores$week <- factor(group$week)
clr_scores$subject <- factor(group$subject)

```
plot
```{r}
# extract the stress value from the NMDS model
stress_value_2 = clr_nmds_2$stress
stress_value_3 = clr_nmds_3$stress

p_NMDS12_clr_treat <- ggplot(data = clr_scores) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = treatment)) + 
  scale_color_manual(values = c( "#332288","#E066A6"),
                     labels = c("Half-cup", "Two cups")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = treatment, color = treatment), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6"),
                    labels = c("Half-cup", "Two cups")) +
  # labs(title= paste0("Euclidean Distances (Stress: ", round(stress_value_2, 2), ")")) +
  theme_pubclean()+
  theme(text = element_text(size = 12),
        plot.title = element_text(face = "bold")) + 
  facet_grid(~week, labeller = as_labeller(c("0" = "Week 0", "8"= "Week 8"))) +
  labs(title = "A", subtitle= "Bacteria",color = "", fill = "")+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "bold", size = 12, hjust = 0.5))
p_NMDS12_clr_treat

```

```{r}
# Putting treatments on the same graph
p_NMDS12_clr_treat_2 <- ggplot(data = clr_scores) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = week)) + 
  scale_color_manual(values = c( "#EC172B","#0DAFB5"),
                     labels = c("Week 0", "Week 8")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = week, color = week), alpha = 0.25) +
  scale_fill_manual(values = c( "#EC172B","#0DAFB5"),
                    labels = c("Week 0", "Week 8")) +
  # labs(title= paste0("Euclidean Distances (Stress: ", round(stress_value_2, 2), ")")) +
  theme_pubclean()+
  theme(text = element_text(size = 12),
        plot.title = element_text(face = "bold")) + 
  facet_grid(~treatment, labeller = as_labeller(c("H" = "Half cup", "T"= "Two cup"))) +
  labs(title = "A", subtitle= "Bacteria",color = "", fill = "")+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "bold", size = 12, hjust = 0.5))

p_NMDS12_clr_treat_2
```
Save plots as .png images
```{r}
ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/NMDS/NMDS12-plot_clr-treat.png", plot = p_NMDS12_clr_treat, width = 6, height = 6, dpi = 300)


```
### 3.1.6 PERMANOVA - stats for beta diversity
Get distances
```{r}
dist_asv <- vegdist(as.matrix(t(asv_tx)), method ='euclidean')

```
do 10000 permutations
```{r}


perm_asv_treat <- adonis2(dist_asv ~ treatment * week, group, permutations = 9999)

# perm_asv <- adonis2(dist_asv ~ treatment * week, group, permutations = 9999, method="euclidean", by="margin", parallel = 4)
perm_asv_treat

```
```{r}
# Add by = terms to get a p-value for each factor
perm_asv_treat_2 <- adonis2(dist_asv ~ treatment * week, data = group,permutations = 9999,by = "terms")


perm_asv_treat_2
```




Save PERMANOVA results as .csv file
```{r}
write.csv(perm_asv_treat, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/PERMANOVA/PERMANOVA-results_asv_clr.csv", row.names = FALSE)

```
### 3.1.6 Beta-dispersion (distance from centroid of group)
Make factor vectors for the data
```{r}
trt_factor <- factor(group$treatment)
week_factor <- factor(group$week)

# Calculate beta dispersion for each grouping - this uses the distance values calculated in the PERMANOVA section
# betadisper() calculates the dispersion for each group, put the distances from betadisper() and 'trt_factor' into a dataframe to graph them
# permutest() is a permutation-based test of multivariate homogeneity of group dispersions; null hypothesis = no difference in dispersion between groups -- if the groups do not have similar variance (if the p value is high), use nonparametric stats tests on them

```

```{r}
trt_betadist_clr <- betadisper(dist_asv, trt_factor)
distances_trt_clr <- data.frame(trt_betadist_clr[['distances']])
distances_trt_clr <- cbind(distances_trt_clr, trt_factor)
colnames(distances_trt_clr) <- c('Distance to Centroid', 'Treatment')
head(distances_trt_clr)

week_betadist_clr <- betadisper(dist_asv, week_factor)
distances_week_clr <- data.frame(week_betadist_clr[['distances']])
distances_week_clr <- cbind(distances_week_clr, week_factor)
colnames(distances_week_clr) <- c('Distance to Centroid', 'Week')
head(distances_week_clr)

```

```{r}
test_trt_betadist_clr <- permutest(trt_betadist_clr)
test_trt_betadist_clr

test_week_betadist_clr <- permutest(week_betadist_clr)
test_week_betadist_clr

# Do stats next -- 

```
### graph beta dispersion
```{r}
B_trt_clr <- ggboxplot(data = distances_trt_clr, x = "Treatment", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
B_trt_clr

B_week_clr <- ggboxplot(data = distances_week_clr, x = "Week", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
B_week_clr

```

```{r}
betadipsr_figure_clr <- ggarrange(B_trt_clr, B_week_clr,
                              labels = c("A", "B"),
                              ncol = 2, nrow = 1)
betadipsr_figure_clr

```


```{r}

# Combine treatment and week into a new factor for use in betadisper
group$treatmentweek <- paste(group$treatment, group$week, sep = " & ")

# Create combined factor
combined_factor <- factor(group$treatmentweek)

# Apply betadisper using the combined factor
combined_betadist_clr <- betadisper(dist_asv, combined_factor)

# Calculate the distances for each combination of Day and Diet
distances_combined_clr <- data.frame(SampleID = rownames(combined_betadist_clr$vectors), 
                                     Distance_to_Centroid = combined_betadist_clr[['distances']])

# Merge with the original group data to get the Day, Diet, and combined factor info
distances_combined_clr <- merge(distances_combined_clr, group[, c("SampleID", "treatment", "week", "treatmentweek")], by = "SampleID", all.x = TRUE)

# View the result
head(distances_combined_clr)



```



```{r}

# Now we can graph
distances_combined_clr$treatmentweek <- factor(distances_combined_clr$treatmentweek)
combined_diet_clr <- ggboxplot(data = distances_combined_clr, x = "treatmentweek", y = "Distance_to_Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
combined_diet_clr



```
## 3.2 Weighted UniFrac Distances
Check dimensions of UniFrac matrix
```{r}
dim(weighted_unfiltered)

# assign matrix to new variable
weighted_unifrac_dist_matrix <- weighted_unfiltered
dim(weighted_unifrac_dist_matrix)

```
filter out samples according to 'group' and complete.cases()
```{r}
weighted_unifrac_dist_matrix <- weighted_unifrac_dist_matrix[(rownames(weighted_unifrac_dist_matrix) %in% group$SampleID),]
weighted_unifrac_dist_matrix <- weighted_unifrac_dist_matrix[,(colnames(weighted_unifrac_dist_matrix) %in% group$SampleID)]

# check dimensions one more time
dim(weighted_unifrac_dist_matrix)

```
### 3.2.1 NMDS
check how many dimensions to use for NMDS
```{r}
w_uni_nmds_2 = metaMDS(weighted_unifrac_dist_matrix, distance = "euclidean", parallel = 2, k = 2)
w_uni_nmds_3 = metaMDS(weighted_unifrac_dist_matrix, distance = "euclidean", parallel = 2, k = 3)

```
extract NMDS scores
```{r}
w_uni_scores = as.data.frame(scores(w_uni_nmds_2))

```
add map data to scores
```{r}
w_uni_scores$treatment <- group$treatment
w_uni_scores$week <- factor(group$week)

```
plot
```{r}
p_NMDS12_weighted_unifrac_treatment <- ggplot(data = w_uni_scores) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = treatment)) + 
  scale_color_manual(values = c( "#332288","#E066A6", "#E6A316", "#256769")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = treatment, color = treatment), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6", "#E6A316", "#256769")) +
  theme_classic() +
  labs(title= "Weighted Unifrac Distances") + # labs(title= "Euclidean Distances")
  theme(text = element_text(size = 15)) + 
  facet_wrap(~week)
p_NMDS12_weighted_unifrac_treatment

p_NMDS12_weighted_unifrac_week <- ggplot(data = w_uni_scores) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = week)) + 
  scale_color_manual(values = c( "#332288","#E066A6", "#E6A316", "#256769")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = week, color = week), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6", "#E6A316", "#256769")) +
  theme_classic() +
  labs(title= "Weighted Unifrac Distances") + # labs(title= "Euclidean Distances")
  theme(text = element_text(size = 15)) + 
  facet_wrap(~treatment)
p_NMDS12_weighted_unifrac_week

```
Save plots as .png images
```{r}
#ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/NMDS/NMDS12-plot_weighted_unifrac.png", plot = p_NMDS12_weighted_unifrac, width = 6, height = 6, dpi = 300)

```
### 3.2.6 PERMANOVA - stats for beta diversity
Make distance object (dist object can be used in NMDS or PCoA)
```{r}
weighted_unifrac_dist <- as.dist(weighted_unifrac_dist_matrix)

```
do 10000 permutations
```{r}
perm_weighted_unifrac <- adonis2(weighted_unifrac_dist ~ treatment * week, group, permutations = 9999, method="euclidean", by="margin", parallel = 4)
perm_weighted_unifrac

```
Save PERMANOVA results as .csv file
```{r}
write.csv(perm_unw_unifrac, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/PERMANOVA/PERMANOVA-results_unweighted_unifrac.csv", row.names = FALSE)

```
### 3.2.6 Beta-dispersion (distance from centroid of group)
Make factor vectors for the data
```{r}
trt_factor <- factor(group$treatment)
week_factor <- factor(group$week)

# Calculate beta dispersion for each grouping - this uses the distance values calculated in the PERMANOVA section
# betadisper() calculates the dispersion for each group, put the distances from betadisper() and 'trt_factor' into a dataframe to graph them
# permutest() is a permutation-based test of multivariate homogeneity of group dispersions; null hypothesis = no difference in dispersion between groups -- if the groups do not have similar variance (if the p value is high), use nonparametric stats tests on them

```

```{r}
trt_betadist_weighted_uni <- betadisper(weighted_unifrac_dist, trt_factor)
distances_trt_weighted_uni <- data.frame(trt_betadist_weighted_uni[['distances']])
distances_trt_weighted_uni <- cbind(distances_trt_weighted_uni, trt_factor)
colnames(distances_trt_weighted_uni) <- c('Distance to Centroid', 'Treatment')
head(distances_trt_weighted_uni)

week_betadist_weighted_uni <- betadisper(weighted_unifrac_dist, week_factor)
distances_week_weighted_uni <- data.frame(trt_betadist_weighted_uni[['distances']])
distances_week_weighted_uni <- cbind(distances_week_weighted_uni, week_factor)
colnames(distances_week_weighted_uni) <- c('Distance to Centroid', 'Week')
head(distances_week_weighted_uni)

```

```{r}
test_trt_betadist_weighted_uni <- permutest(trt_betadist_weighted_uni)
test_trt_betadist_weighted_uni

test_week_betadist_weighted_uni <- permutest(week_betadist_weighted_uni)
test_week_betadist_weighted_uni

# Do stats next -- 

```
### graph beta dispersion
```{r}
B_trt_weighted_uni <- ggboxplot(data = distances_trt_weighted_uni, x = "Treatment", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "UniFrac Distances")
B_trt_weighted_uni

B_week_weighted_uni <- ggboxplot(data = distances_week_weighted_uni, x = "Week", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "UniFrac Distances")
B_week_weighted_uni

```

```{r}
betadipsr_figure_weighted_uni <- ggarrange(B_trt_weighted_uni, B_week_weighted_uni,
                              labels = c("A", "B"),
                              ncol = 2, nrow = 1)
betadipsr_figure_weighted_uni

```
```{r}
# Calc betadisper using the combined factor
combined_betadist_weighted_uni <- betadisper(weighted_unifrac_dist, combined_factor)

# Calc the distances for each combination of Day and Diet
distances_combined_weighted_uni <- data.frame(SampleID = rownames(combined_betadist_weighted_uni$vectors), 
                                     Distance_to_Centroid = combined_betadist_weighted_uni[['distances']])

# Merge with group to include treatment, week, and treatmentweek info
distances_combined_weighted_uni <- merge(distances_combined_weighted_uni, group[, c("SampleID", "treatment", "week", "treatmentweek")], 
                                by = "SampleID", all.x = TRUE)
# Double check
head(distances_combined_weighted_uni)


```

# Now we can graph
```{r}
distances_combined_weighted_uni$treatmentweek <- factor(distances_combined_weighted_uni$treatmentweek)
combined_treatment_weighted_uni <- ggboxplot(data = distances_combined_weighted_uni, x = "treatmentweek", y = "Distance_to_Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Weighted Unifrac Distances") +
scale_color_manual(values = c("H & 0" = "#17AB0E", 
                              "T & 0" = "#E22E8C", 
                              "H & 8" = "#17AB0E", 
                              "T & 8" = "#E22E8C")) 


combined_treatment_weighted_uni


combined_treatment_weighted_uni <- ggboxplot(data = distances_combined_weighted_uni, x = "treatmentweek", y = "Distance_to_Centroid") +
  theme_classic() +
  geom_boxplot(aes(fill = treatmentweek), alpha = 0.5, color = "black") +  
  scale_fill_manual(values = c("H & 0" = "#17AB0E", 
                               "H & 8" = "#17AB0E", 
                               "T & 0" = "#E22E8C", 
                               "T & 8" = "#E22E8C")) +  
  labs(title = "Weighted Unifrac Distances",
       x = "Treatment & Week",        
       y = "Beta Dispersion") 

combined_treatment_weighted_uni


```

## 3.3 Unweighted UniFrac Distances
Check dimensions of UniFrac matrix
```{r}
# use unweighted_unfiltered
dim(unweighted_unfiltered)

# assign matrix to new variable
unweighted_unifrac_dist_matrix <- unweighted_unfiltered
dim(unweighted_unifrac_dist_matrix)

```

```{r}
unweighted_unifrac_dist_matrix <- unweighted_unifrac_dist_matrix[(rownames(unweighted_unifrac_dist_matrix) %in% group$SampleID),]
unweighted_unifrac_dist_matrix <- unweighted_unifrac_dist_matrix[,(colnames(unweighted_unifrac_dist_matrix) %in% group$SampleID)]

# check dimensions one more time
dim(unweighted_unifrac_dist_matrix)

```
### 3.3.1 NMDS
check how many dimensions to use for NMDS
```{r}
unw_uni_nmds_2 = metaMDS(unweighted_unifrac_dist_matrix, distance = "euclidean", parallel = 2, k = 2)
unw_uni_nmds_3 = metaMDS(unweighted_unifrac_dist_matrix, distance = "euclidean", parallel = 2, k = 3)

```

extract NMDS scores
```{r}
unw_uni_scores = as.data.frame(scores(unw_uni_nmds_2))

```
add map data to scores
```{r}
unw_uni_scores$treatment <- group$treatment
unw_uni_scores$week <- factor(group$week)

```

plot
```{r}
p_NMDS12_unweighted_unifrac_treatment <- ggplot(data = unw_uni_scores) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = treatment)) + 
  scale_color_manual(values = c( "#332288","#E066A6")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = treatment, color = treatment), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6")) +
  theme_classic() +
  labs(title= "Unweighted Unifrac Distances") + # labs(title= "Euclidean Distances")
  theme(text = element_text(size = 15)) + 
  facet_wrap(~week)
p_NMDS12_unweighted_unifrac_treatment

p_NMDS12_unweighted_unifrac_week <- ggplot(data = unw_uni_scores) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = week)) + 
  scale_color_manual(values = c( "#332288","#E066A6")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = week, color = week), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6")) +
  theme_classic() +
  labs(title= "Unweighted Unifrac Distances") + # labs(title= "Euclidean Distances")
  theme(text = element_text(size = 15)) + 
  facet_wrap(~treatment)
p_NMDS12_unweighted_unifrac_week

```
### 3.3.6 PERMANOVA - stats for beta diversity
Make distance object (dist object can be used in NMDS or PCoA)
```{r}
unweighted_unifrac_dist <- as.dist(unweighted_unifrac_dist_matrix)

```
do 10000 permutations
```{r}
perm_unweighted_unifrac <- adonis2(unweighted_unifrac_dist ~ treatment * week, group, permutations = 9999, method="euclidean", by="margin", parallel = 4)
perm_unweighted_unifrac

```
Save PERMANOVA results as .csv file
```{r}
write.csv(perm_unw_unifrac, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/PERMANOVA/PERMANOVA-results_unweighted_unifrac.csv", row.names = FALSE)

```
### 3.3.6 Beta-dispersion (distance from centroid of group)
Make factor vectors for the data
```{r}
trt_factor <- factor(group$treatment)
week_factor <- factor(group$week)

# Calculate beta dispersion for each grouping - this uses the distance values calculated in the PERMANOVA section
# betadisper() calculates the dispersion for each group, put the distances from betadisper() and 'trt_factor' into a dataframe to graph them
# permutest() is a permutation-based test of multivariate homogeneity of group dispersions; null hypothesis = no difference in dispersion between groups -- if the groups do not have similar variance (if the p value is high), use nonparametric stats tests on them

```

```{r}
trt_betadist_unweighted_uni <- betadisper(unweighted_unifrac_dist, trt_factor)
distances_trt_unweighted_uni <- data.frame(trt_betadist_unweighted_uni[['distances']])
distances_trt_unweighted_uni <- cbind(distances_trt_unweighted_uni, trt_factor)
colnames(distances_trt_unweighted_uni) <- c('Distance to Centroid', 'Treatment')
head(distances_trt_unweighted_uni)

week_betadist_unweighted_uni <- betadisper(unweighted_unifrac_dist, week_factor)
distances_week_unweighted_uni <- data.frame(trt_betadist_unweighted_uni[['distances']])
distances_week_unweighted_uni <- cbind(distances_week_unweighted_uni, week_factor)
colnames(distances_week_unweighted_uni) <- c('Distance to Centroid', 'Week')
head(distances_week_unweighted_uni)

```
```{r}
test_trt_betadist_unweighted_uni <- permutest(trt_betadist_unweighted_uni)
test_trt_betadist_unweighted_uni

test_week_betadist_unweighted_uni <- permutest(week_betadist_unweighted_uni)
test_week_betadist_unweighted_uni

# Do stats next -- 

```

### graph beta dispersion
```{r}
B_trt_unweighted_uni <- ggboxplot(data = distances_trt_unweighted_uni, x = "Treatment", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Unweighted UniFrac Distances")
B_trt_unweighted_uni

B_week_unweighted_uni <- ggboxplot(data = distances_week_unweighted_uni, x = "Week", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Unweighted UniFrac Distances")
B_week_unweighted_uni

```
```{r}
betadipsr_figure_unweighted_uni <- ggarrange(B_trt_unweighted_uni, B_week_unweighted_uni,
                              labels = c("A", "B"),
                              ncol = 2, nrow = 1)
betadipsr_figure_unweighted_uni

```

```{r}
# Calc betadisper using the combined factor
combined_betadist_unweighted_uni <- betadisper(unweighted_unifrac_dist, combined_factor)

# Calc the distances for each combination of Day and Diet
distances_combined_unweighted_uni <- data.frame(SampleID = rownames(combined_betadist_unweighted_uni$vectors), 
                                     Distance_to_Centroid = combined_betadist_unweighted_uni[['distances']])

# Merge with group to include treatment, week, and treatmentweek info
distances_combined_unweighted_uni <- merge(distances_combined_unweighted_uni, group[, c("SampleID", "treatment", "week", "treatmentweek")], 
                                by = "SampleID", all.x = TRUE)
# Double check
head(distances_combined_unweighted_uni)


```
# Now we can graph
```{r}
distances_combined_unweighted_uni$treatmentweek <- factor(distances_combined_unweighted_uni$treatmentweek)
combined_treatment_unweighted_uni <- ggboxplot(data = distances_combined_unweighted_uni, x = "treatmentweek", y = "Distance_to_Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Unweighted Unifrac Distances") +
scale_color_manual(values = c("H & 0" = "#17AB0E", 
                              "T & 0" = "#E22E8C", 
                              "H & 8" = "#17AB0E", 
                              "T & 8" = "#E22E8C")) 


combined_treatment_unweighted_uni


combined_treatment_unweighted_uni <- ggboxplot(data = distances_combined_unweighted_uni, x = "treatmentweek", y = "Distance_to_Centroid") +
  theme_classic() +
  geom_boxplot(aes(fill = treatmentweek), alpha = 0.5, color = "black") +  
  scale_fill_manual(values = c("H & 0" = "#17AB0E", 
                               "H & 8" = "#17AB0E", 
                               "T & 0" = "#E22E8C", 
                               "T & 8" = "#E22E8C")) +  
  labs(title = "Unweighted Unifrac Distances",
       x = "Treatment & Week",        
       y = "Beta Dispersion") 

combined_treatment_unweighted_uni


```




### Saving clr transformed data for PyBootNet

```{r}


transposed_clr_asv_save <- transposed_clr_asv
SampleIDs <- rownames(transposed_clr_asv_save)
rownames(transposed_clr_asv_save) <- NULL
transposed_clr_asv_save <- cbind(SampleID,transposed_clr_asv_save)


write.csv(transposed_clr_asv_save,"transformed_asv.csv", row.names = FALSE)


```


```{r}


transposed_clr_asv_20 <- data.frame(t(transformed_filtered_asv_20))
transposed_clr_asv_20_save <- transposed_clr_asv_20
SampleIDs <- rownames(transposed_clr_asv_20_save)
rownames(transposed_clr_asv_20_save) <- NULL
transposed_clr_asv_20_save <- cbind(SampleID,transposed_clr_asv_save)


write.csv(transposed_clr_asv_20_save,"transformed_filtered20_asv.csv", row.names = FALSE)


```
