---
title: "Microbiome Fruit Metabolites - Beta Diversity"
author: "Shyan Polman"
date: "`r Sys.Date()`" 
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
library(vegan)
library(ggplot2)
library(ggpubr)


```

# Import the fecal metabolite csv

```{r}

fecal_metabolites <- read.csv("~/microbiome-fruit/Hooshmand_LCMP_Feces_Metabolites.csv", header = TRUE) 

# Replace the "." back to "-" in the Sample IDs
colnames(fecal_metabolites) <- gsub("\\.", "-", colnames(fecal_metabolites))


# Use values from first column as row names
rownames(fecal_metabolites) <- fecal_metabolites[,1]

# Remove first column from data frame
fecal_metabolites <- fecal_metabolites[,-1]

# Replace NAs with 0

fecal_metabolites[is.na(fecal_metabolites)] <- 0

head(fecal_metabolites)



# Data is now formatted to proceed with beta diversity analysis
 
```

```{r}


transposed_fecal = as.data.frame(t(fecal_metabolites))
write.csv(transposed_fecal, "transposed_fecal.csv")
get.cwd()


```

# 3. Beta Diversity
## 3.1 CLR transformation
### 3.1.1 Replace zeros with pseudocount 0 = 0.001
```{r}
fecal_metabolites_zerorep <- fecal_metabolites
fecal_metabolites_zerorep[fecal_metabolites_zerorep == 0] <- 0.001
# tail(asv_zerorep)

```
### 3.1.2 Create function with clr formula to apply transformation (for clr formula, samples = columns)
```{r}
clr <- function(x) sweep(log(x), 1, rowMeans(log(x)), "-")

# log(x): calculates the natural logarithm of each element in the matrix x; log() computes the natural logarithm
# rowMeans(log(x)): computes the mean of each row in the matrix log(x). rowMeans() calculates the row-wise means
# sweep(log(x), 1, rowMeans(log(x)), "-"): This is where the sweep() function is used - subtracts the row-wise means (obtained from step 2) from each element in the matrix log(x) along the first dimension (rows) - the 1 parameter specifies the dimension along which the operation should be performed, and "-" indicates subtraction

```
### 3.1.3 Apply clr-transformation function to zero-replaced tables
```{r}
transformed_fecal_metabolites <- data.frame(t(clr(t(fecal_metabolites_zerorep))), check.names = F)

```
### 3.1.4 Subset feature data into Treatment or Day groups if needed
Assign 'group' with desired subgroup variable (see microbiome-ocp.Rmd)
```{r}

group <- map_all[,c("SampleID","subject","treatment","week")]
head(group)

```

```{r}
# subset clr-transformed data according to 'group', if needed 

# don't need to do because the dataset does not need subsetting 
# asv_tx <- asv_tx[, colnames(asv_tx) %in% group$SampleID]

```
### 3.1.5 NMDS
transpose data frames (samples = rows) and make matrix
```{r}
clr_fecalmet_matrix <- as.matrix(t(transformed_fecal_metabolites))

```
check how many dimensions to use for NMDS
```{r}
clr_fecalmet_matrix_nmds_2 = metaMDS(clr_fecalmet_matrix, distance = "euclidean", parallel = 2, k = 2)
clr_fecalmet_matrix_nmds_3 = metaMDS(clr_fecalmet_matrix, distance = "euclidean", parallel = 2, k = 3)
clr_fecalmet_matrix_nmds_2$stress
clr_fecalmet_matrix_nmds_3$stress
```
extract NMDS scores
```{r}
clr_scores_fecalm = as.data.frame(scores(clr_fecalmet_matrix_nmds_3))

```
add map data to scores
```{r}
clr_scores_fecalm$treatment <- group$treatment
clr_scores_fecalm$week <- factor(group$week)
clr_scores_fecalm$subject <- factor(group$subject)

```
plot
```{r}
# extract the stress value from the NMDS model
clr_scores_fecalm$week = factor(clr_scores_fecalm$week)
stress_value_2_fecalm = clr_fecalmet_matrix_nmds_2$stress
stress_value_3_fecalm = clr_fecalmet_matrix_nmds_3$stress

p_NMDS12_clr_fecalm_treat <- ggplot(data = clr_scores_fecalm) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = treatment)) + 
  scale_color_manual(values = c( "#332288","#E066A6"),
                     labels = c("Half-cup", "Two cups")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = treatment, color = treatment), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6"),
                    labels = c("Half-cup", "Two cups")) +
  # labs(title= paste0("Euclidean Distances (Stress: ", round(stress_value_2, 2), ")")) +
  theme_pubclean()+
  theme(text = element_text(size = 12),
        plot.title = element_text(face = "bold")) + 
  facet_grid(~week, labeller = as_labeller(c("0" = "Week 0", "8"= "Week 8"))) +
  labs(title = "B", subtitle= "Fecal Metabolites",color = "", fill = "")+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "bold", size = 12, hjust = 0.5))
p_NMDS12_clr_fecalm_treat

```
```{r}

# extract the stress value from the NMDS model
clr_scores_fecalm$week = factor(clr_scores_fecalm$week)
stress_value_2_fecalm = clr_fecalmet_matrix_nmds_2$stress
stress_value_3_fecalm = clr_fecalmet_matrix_nmds_3$stress

p_NMDS12_clr_fecalm_treat_2 <- ggplot(data = clr_scores_fecalm) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = week)) + 
  scale_color_manual(values = c( "#EC172B","#0DAFB5"),
                     labels = c("Week 0", "Week 8")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = week, color = week), alpha = 0.25) +
  scale_fill_manual(values = c( "#EC172B","#0DAFB5"),
                    labels = c("Week 0", "Week 8")) +
  # labs(title= paste0("Euclidean Distances (Stress: ", round(stress_value_2, 2), ")")) +
  theme_pubclean()+
  theme(text = element_text(size = 12),
        plot.title = element_text(face = "bold")) + 
  facet_grid(~treatment, labeller = as_labeller(c("H" = "Half cup", "T"= "Two cup"))) +
  labs(title = "B", subtitle= "Fecal Metabolites",color = "", fill = "")+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "bold", size = 12, hjust = 0.5))
p_NMDS12_clr_fecalm_treat_2

```
Save plots as .png images
```{r}
#ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/NMDS/NMDS12-plot_clr-treat.png", plot = p_NMDS12_clr_treat, width = 6, height = 6, dpi = 300)


```
### 3.1.6 PERMANOVA - stats for beta diversity
Get distances
```{r}
dist_fecal_metabolites <- vegdist(as.matrix(t(transformed_fecal_metabolites)), method ='euclidean')

```
do 10000 permutations
```{r}
perm_fecal_metabolites <- adonis2(dist_fecal_metabolites ~ treatment * week, group, permutations = 9999, method="euclidean", by="terms")

# perm_asv <- adonis2(dist_asv ~ treatment * week, group, permutations = 9999, method="euclidean", by="margin", parallel = 4)
perm_fecal_metabolites

```
Save PERMANOVA results as .csv file
```{r}
write.csv(perm_asv_treat, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/PERMANOVA/PERMANOVA-results_asv_clr.csv", row.names = FALSE)

```
### 3.1.6 Beta-dispersion (distance from centroid of group)
Make factor vectors for the data
```{r}
trt_factor <- factor(group$treatment)
week_factor <- factor(group$week)

# Calculate beta dispersion for each grouping - this uses the distance values calculated in the PERMANOVA section
# betadisper() calculates the dispersion for each group, put the distances from betadisper() and 'trt_factor' into a dataframe to graph them
# permutest() is a permutation-based test of multivariate homogeneity of group dispersions; null hypothesis = no difference in dispersion between groups -- if the groups do not have similar variance (if the p value is high), use nonparametric stats tests on them

```

```{r}
fecalm_trt_betadist_clr <- betadisper(dist_fecal_metabolites, trt_factor)
fecalm_distances_trt_clr <- data.frame(fecalm_trt_betadist_clr[['distances']])
fecalm_distances_trt_clr <- cbind(fecalm_distances_trt_clr, trt_factor)
colnames(fecalm_distances_trt_clr) <- c('Distance to Centroid', 'Treatment')
head(fecalm_distances_trt_clr)

fecalm_week_betadist_clr <- betadisper(dist_fecal_metabolites, week_factor)
fecalm_distances_week_clr <- data.frame(fecalm_week_betadist_clr[['distances']])
fecalm_distances_week_clr <- cbind(fecalm_distances_week_clr, week_factor)
colnames(fecalm_distances_week_clr) <- c('Distance to Centroid', 'Week')
head(fecalm_distances_week_clr)

```

```{r}
fecalm_test_trt_betadist_clr <- permutest(fecalm_trt_betadist_clr)
fecalm_test_trt_betadist_clr

fecalm_test_week_betadist_clr <- permutest(fecalm_week_betadist_clr)
fecalm_test_week_betadist_clr

# Do stats next -- 

```
### graph beta dispersion
```{r}
fecalm_B_trt_clr <- ggboxplot(data = fecalm_distances_trt_clr, x = "Treatment", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
fecalm_B_trt_clr

fecalm_B_week_clr <- ggboxplot(data = fecalm_distances_week_clr, x = "Week", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
fecalm_B_week_clr

```

```{r}
fecalm_betadipsr_figure_clr <- ggarrange(fecalm_B_trt_clr, fecalm_B_week_clr,
                              labels = c("A", "B"),
                              ncol = 2, nrow = 1)
fecalm_betadipsr_figure_clr

```



```{r}

# Combine treatment and week into a new factor for use in betadisper
group$treatmentweek <- paste(group$treatment, group$week, sep = " & ")

# Create combined factor
combined_factor <- factor(group$treatmentweek)

# Apply betadisper using the combined factor
fecalm_combined_betadist_clr <- betadisper(dist_fecal_metabolites, combined_factor)

# Calculate the distances for each combination of Day and Diet
fecalm_distances_combined_clr <- data.frame(SampleID = rownames(fecalm_combined_betadist_clr$vectors), 
                                     Distance_to_Centroid = fecalm_combined_betadist_clr[['distances']])

# Merge with the original group data to get the Day, Diet, and combined factor info
fecalm_distances_combined_clr <- merge(fecalm_distances_combined_clr, group[, c("SampleID", "treatment", "week", "treatmentweek")], by = "SampleID", all.x = TRUE)

# View the result
head(fecalm_distances_combined_clr)



```




```{r}

# Now we can graph
fecalm_distances_combined_clr$treatmentweek <- factor(fecalm_distances_combined_clr$treatmentweek)
fecalm_combined_diet_clr <- ggboxplot(data = fecalm_distances_combined_clr, x = "treatmentweek", y = "Distance_to_Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
fecalm_combined_diet_clr



```




# Import the plasma metabolite csv

```{r}

plasma_metabolites <- read.csv("~/microbiome-fruit/Hooshmand_LCMP_Plasma_Metabolites.csv", header = TRUE) 

# Replace the "." back to "-" in the Sample IDs
colnames(plasma_metabolites) <- gsub("\\.", "-", colnames(plasma_metabolites))


# Use values from first column as row names
rownames(plasma_metabolites) <- plasma_metabolites[,1]

# Remove first column from data frame
plasma_metabolites <- plasma_metabolites[,-1]

# Replace NAs with 0

plasma_metabolites[is.na(plasma_metabolites)] <- 0

head(plasma_metabolites)

# Data is now formatted to proceed with beta diversity analysis
 
```

```{r}

transposed_plasma = as.data.frame(t(plasma_metabolites))
write.csv(transposed_plasma, "transposed_plasma.csv")

```



## 3.2 CLR transformation
### 3.2.1 Replace zeros with pseudocount 0 = 0.001
```{r}
plasma_metabolites_zerorep <- plasma_metabolites
plasma_metabolites_zerorep[plasma_metabolites_zerorep == 0] <- 0.001
# tail(asv_zerorep)

```
### 3.2.2 Create function with clr formula to apply transformation (for clr formula, samples = columns)
```{r}
clr <- function(x) sweep(log(x), 1, rowMeans(log(x)), "-")

# log(x): calculates the natural logarithm of each element in the matrix x; log() computes the natural logarithm
# rowMeans(log(x)): computes the mean of each row in the matrix log(x). rowMeans() calculates the row-wise means
# sweep(log(x), 1, rowMeans(log(x)), "-"): This is where the sweep() function is used - subtracts the row-wise means (obtained from step 2) from each element in the matrix log(x) along the first dimension (rows) - the 1 parameter specifies the dimension along which the operation should be performed, and "-" indicates subtraction

```
### 3.2.3 Apply clr-transformation function to zero-replaced tables
```{r}
transformed_plasma_metabolites <- data.frame(t(clr(t(plasma_metabolites_zerorep))), check.names = F)

```
### 3.2.4 Subset feature data into Treatment or Day groups if needed
Assign 'group' with desired subgroup variable (see microbiome-ocp.Rmd)
```{r}

group <- map_all[,c("SampleID","subject","treatment","week")]
head(group)

```

```{r}
# subset clr-transformed data according to 'group', if needed 

# don't need to do because the dataset does not need subsetting 
# asv_tx <- asv_tx[, colnames(asv_tx) %in% group$SampleID]

```
### 3.2.5 NMDS
transpose data frames (samples = rows) and make matrix
```{r}
clr_plasmamet_matrix <- as.matrix(t(transformed_plasma_metabolites))

```
check how many dimensions to use for NMDS
```{r}
clr_plasmamet_matrix_nmds_2 = metaMDS(clr_plasmamet_matrix, distance = "euclidean", parallel = 2, k = 2)
clr_plasmamet_matrix_nmds_3 = metaMDS(clr_plasmamet_matrix, distance = "euclidean", parallel = 2, k = 3)
clr_plasmamet_matrix_nmds_2$stress
clr_plasmamet_matrix_nmds_3$stress
```
extract NMDS scores
```{r}
clr_scores_plasmam = as.data.frame(scores(clr_plasmamet_matrix_nmds_3))

```
add map data to scores
```{r}
clr_scores_plasmam$treatment <- group$treatment
clr_scores_plasmam$week <- factor(group$week)
clr_scores_plasmam$subject <- factor(group$subject)

```
plot
```{r}
# extract the stress value from the NMDS model
stress_value_2_plasmam = clr_plasmamet_matrix_nmds_2$stress
stress_value_3_plasmam = clr_plasmamet_matrix_nmds_3$stress

p_NMDS12_clr_plasmam_treat <- ggplot(data = clr_scores_plasmam) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = treatment)) + 
  scale_color_manual(values = c( "#332288","#E066A6"),
                     labels = c("Half-cup", "Two cups")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = treatment, color = treatment), alpha = 0.25) +
  scale_fill_manual(values = c( "#332288","#E066A6"),
                    labels = c("Half-cup", "Two cups")) +
  # labs(title= paste0("Euclidean Distances (Stress: ", round(stress_value_2, 2), ")")) +
  theme_pubclean()+
  theme(text = element_text(size = 12),
        plot.title = element_text(face = "bold")) + 
  facet_grid(~week, labeller = as_labeller(c("0" = "Week 0", "8"= "Week 8"))) +
  labs(title = "C", subtitle= "Plasma Metabolites",color = "", fill = "")+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "bold", size = 12, hjust = 0.5))

p_NMDS12_clr_plasmam_treat

```
```{r}
# extract the stress value from the NMDS model
clr_scores_plasmam$week = factor(clr_scores_plasmam$week)
stress_value_2_plasmam = clr_plasmamet_matrix_nmds_2$stress
stress_value_3_plasmam = clr_plasmamet_matrix_nmds_3$stress

p_NMDS12_clr_plasmam_treat_2 <- ggplot(data = clr_scores_plasmam) +
  geom_point(aes(x = NMDS1, y = NMDS2, color = week)) + 
  scale_color_manual(values = c( "#EC172B","#0DAFB5"),
                     labels = c("Week 0", "Week 8")) +
  stat_ellipse(geom = "polygon",aes(x = NMDS1, y = NMDS2, fill = week, color = week), alpha = 0.25) +
  scale_fill_manual(values = c( "#EC172B","#0DAFB5"),
                    labels = c("Week 0", "Week 8")) +
  # labs(title= paste0("Euclidean Distances (Stress: ", round(stress_value_2, 2), ")")) +
  theme_pubclean()+
  theme(text = element_text(size = 12),
        plot.title = element_text(face = "bold")) + 
  facet_grid(~treatment, labeller = as_labeller(c("H" = "Half cup", "T"= "Two cup"))) +
  labs(title = "C", subtitle= "Plasma Metabolites",color = "", fill = "")+
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0),
    plot.subtitle = element_text(face = "bold", size = 12, hjust = 0.5))
p_NMDS12_clr_plasmam_treat_2

```



Save plots as .png images
```{r}
#ggsave(filename = "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/NMDS/NMDS12-plot_clr-treat.png", plot = p_NMDS12_clr_treat, width = 6, height = 6, dpi = 300)


```
### 3.2.6 PERMANOVA - stats for beta diversity
Get distances
```{r}
dist_plasma_metabolites <- vegdist(as.matrix(t(transformed_plasma_metabolites)), method ='euclidean')

```
do 10000 permutations
```{r}
perm_plasma_metabolites <- adonis2(dist_plasma_metabolites ~ treatment * week, group, permutations = 9999, method="euclidean", by="terms")

# perm_asv <- adonis2(dist_asv ~ treatment * week, group, permutations = 9999, method="euclidean", by="margin", parallel = 4)
perm_plasma_metabolites

```
Save PERMANOVA results as .csv file
```{r}
write.csv(perm_asv_treat, "~/Documents/Programming/Microbiome/microbiome-ocp/data-analyses/results/beta/PERMANOVA/PERMANOVA-results_asv_clr.csv", row.names = FALSE)

```
### 3.2.6 Beta-dispersion (distance from centroid of group)
Make factor vectors for the data
```{r}
trt_factor <- factor(group$treatment)
week_factor <- factor(group$week)

# Calculate beta dispersion for each grouping - this uses the distance values calculated in the PERMANOVA section
# betadisper() calculates the dispersion for each group, put the distances from betadisper() and 'trt_factor' into a dataframe to graph them
# permutest() is a permutation-based test of multivariate homogeneity of group dispersions; null hypothesis = no difference in dispersion between groups -- if the groups do not have similar variance (if the p value is high), use nonparametric stats tests on them

```

```{r}
plasmam_trt_betadist_clr <- betadisper(dist_plasma_metabolites, trt_factor)
plasmam_distances_trt_clr <- data.frame(plasmam_trt_betadist_clr[['distances']])
plasmam_distances_trt_clr <- cbind(plasmam_distances_trt_clr, trt_factor)
colnames(plasmam_distances_trt_clr) <- c('Distance to Centroid', 'Treatment')
head(plasmam_distances_trt_clr)

plasmam_week_betadist_clr <- betadisper(dist_plasma_metabolites, week_factor)
plasmam_distances_week_clr <- data.frame(plasmam_week_betadist_clr[['distances']])
plasmam_distances_week_clr <- cbind(plasmam_distances_week_clr, week_factor)
colnames(plasmam_distances_week_clr) <- c('Distance to Centroid', 'Week')
head(plasmam_distances_week_clr)

```

```{r}
plasmam_test_trt_betadist_clr <- permutest(plasmam_trt_betadist_clr)
plasmam_test_trt_betadist_clr

plasmam_test_week_betadist_clr <- permutest(plasmam_week_betadist_clr)
plasmam_test_week_betadist_clr

# Do stats next -- 

```
### graph beta dispersion
```{r}
plasmam_B_trt_clr <- ggboxplot(data = plasmam_distances_trt_clr, x = "Treatment", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
plasmam_B_trt_clr

plasmam_B_week_clr <- ggboxplot(data = plasmam_distances_week_clr, x = "Week", y = "Distance to Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
plasmam_B_week_clr

```

```{r}
plasmam_betadipsr_figure_clr <- ggarrange(plasmam_B_trt_clr, plasmam_B_week_clr,
                              labels = c("A", "B"),
                              ncol = 2, nrow = 1)
plasmam_betadipsr_figure_clr

```



```{r}

# Combine treatment and week into a new factor for use in betadisper
group$treatmentweek <- paste(group$treatment, group$week, sep = " & ")

# Create combined factor
combined_factor <- factor(group$treatmentweek)

# Apply betadisper using the combined factor
plasmam_combined_betadist_clr <- betadisper(dist_plasma_metabolites, combined_factor)

# Calculate the distances for each combination of Day and Diet
plasmam_distances_combined_clr <- data.frame(SampleID = rownames(plasmam_combined_betadist_clr$vectors), 
                                     Distance_to_Centroid = plasmam_combined_betadist_clr[['distances']])

# Merge with the original group data to get the Day, Diet, and combined factor info
plasmam_distances_combined_clr <- merge(plasmam_distances_combined_clr, group[, c("SampleID", "treatment", "week", "treatmentweek")], by = "SampleID", all.x = TRUE)

# View the result
head(plasmam_distances_combined_clr)



```




```{r}

# Now we can graph
plasmam_distances_combined_clr$treatmentweek <- factor(plasmam_distances_combined_clr$treatmentweek)
plasmam_combined_diet_clr <- ggboxplot(data = plasmam_distances_combined_clr, x = "treatmentweek", y = "Distance_to_Centroid" ) +
theme_classic() +
geom_point() +
labs(title = "Euclidean Distances")
plasmam_combined_diet_clr



```





### Saving clr transformed data for PyBootNet

```{r}

transposed_clr_fecal_metabolites <- data.frame(clr(t(fecal_metabolites_zerorep)), check.names = F)

transposed_clr_fm_save <- transposed_clr_fecal_metabolites
SampleIDs <- rownames(transposed_clr_fm_save)
rownames(transposed_clr_fm_save) <- NULL
transposed_clr_fm_save <- cbind(SampleID,transposed_clr_fm_save)


write.csv(transposed_clr_fm_save,"transformed_fecal_metabolites.csv", row.names = FALSE)



transposed_clr_plasma_metabolites <- data.frame(clr(t(plasma_metabolites_zerorep)), check.names = F)

transposed_clr_plasma_save <- transposed_clr_plasma_metabolites
SampleIDs <- rownames(transposed_clr_plasma_save)
rownames(transposed_clr_plasma_save) <- NULL
transposed_clr_plasma_save <- cbind(SampleID,transposed_clr_plasma_save)


write.csv(transposed_clr_plasma_save,"transformed_plasma_metabolites.csv", row.names = FALSE)



```

