---
title: "coda4microbiome-fruit for metabolites"
author: "Shyan Polman"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
Set working directory
```{r}

setwd("/Users/shyanpolman/microbiome-fruit/data-analyses")
getwd()

```
Load coda4microbiome package
```{r Load Packages}
library(coda4microbiome)
library(ggplot2)

```

```{r}
# to reproduce the results
set.seed(123)

```
# 5. coda4microbiome - CH4 & H2 levels signatures in Control/Users
Assign "group" with metadata subset

I already established a group in microbiome-fruit_pub.Rmd [insert group code below]
```{r}

fecal_metabolites <- read.csv("~/microbiome-fruit/Hooshmand_LCMP_Feces_Metabolites.csv", header = TRUE)

colnames(fecal_metabolites)[1] <- 'Metabolite Name'


# Replace the "." back to "-" in the Sample IDs
colnames(fecal_metabolites) <- gsub("\\.", "-", colnames(fecal_metabolites))



```

### Run coda4microbiome algorithm

### 1. Create a Transposed Data Frame. A transposed data frame x_species is created from the original data frame species

### 2. Factorize the half-cup/two-cup treatment groups. Create a factor variable y vector  by extracting treatment information for 'Half-cup' & 'Two-cup' subjects from the map_all data.

```{r}

#1


x_fecal_metabolites <- data.frame(t(fecal_metabolites), check.names = T)

# Replace the "." back to "-" in the Metabolite Name
colnames(x_fecal_metabolites) <- gsub("\\.", "-", colnames(x_fecal_metabolites))

# Reorder x_species rows to match map_all SampleID order
x_fecal_metabolites_ordered <- x_fecal_metabolites[map_all$SampleID, ]

x_time = map_all$week;

subject_id = map_all$SampleID;

#2 
y = factor(map_all$treatment);

ini_time = 0;

end_time = 8;


# checking the levels

# levels(factor(map_all$treatment))

# Half-cup is the baseline 



```

### 3. Apply regression. Regression is applied using the coda_glmnet_longitudinal function to identify important taxa in 'Half-cup' & 'Two-cup' treatments.
```{r}

set.seed(123)



coda_glmnet_long_fecal <-coda_glmnet_longitudinal(x_fecal_metabolites,y, x_time, subject_id, ini_time, end_time, lambda="lambda.min",nfolds=5)



```


### 4. Extract Accuracy Measure. The apparent Area Under the Curve (AUC) is extracted as an accuracy measure from the regression results.

```{r}
coda_glmnet_long_fecal$taxa.num


coda_glmnet_long_fecal$taxa.name


coda_glmnet_long_fecal$`log-contrast coefficients`



```





```{r}

coda_glmnet_long_fecal$`apparent AUC`
coda_glmnet_long_fecal$`predictions plot`
coda_glmnet_long_fecal$`signature plot`


```






### 5. Create a Data Frame with Genera and Coefficients. A data frame coda_balance_coeff_control_ocis created to store genera names and their corresponding coefficients from the regression
```{r}

coda_balance_coeff_fecal <- data.frame(cbind(coda_glmnet_long_fecal$`log-contrast coefficients`, coda_glmnet_long_fecal$taxa.name))
# view(coda_balance_coeff)

```


### 6. Data Frame Manipulations

```{r}
# Set row names of the data frame coda_balance the taxa names obtained from the regression results
rownames(coda_balance_coeff_fecal) <- coda_balance_coeff_fecal$taxa.name
# Set column names of the data frame to "Coefficient_full" and "Metabolites" for clarity.
colnames(coda_balance_coeff_fecal) <- c("Coefficient_full", "Metabolites")
# Convert "Coefficient_full" column to numeric type, and create a new column "Coefficient" by rounding "Coefficient_full" to two decimal places
coda_balance_coeff_fecal$Coefficient_full <- as.numeric(coda_balance_coeff_fecal$Coefficient_full)
coda_balance_coeff_fecal$Coefficient <-as.numeric(format(round(coda_balance_coeff_fecal$Coefficient_full, 2), nsmall = 2))
# Determine Positive or Negative Coefficients. Create "PosNeg" column based on whether the coefficient is positive or negative; contains the labels "Positive" or "Negative" accordingly.
coda_balance_coeff_fecal$PosNeg <- with(coda_balance_coeff_fecal, ifelse(Coefficient > 0, "Positive", "Negative"))
# Assign Group Labels. The "Group" column is assigned labels "High" or "Low" based on the "PosNeg" column. If the coefficient is positive, it is labeled as "Low"; otherwise, it is labeled as "High".
coda_balance_coeff_fecal$Group <- with(coda_balance_coeff_fecal, ifelse(PosNeg == "Positive", "Low", "High"))
coda_balance_coeff_fecal
```

### 7. Create Data Frame with Samples, Predictions, and Variables. A data frame coda_balances_long_predictions is created to store balance predictions along with sex, group, and sample type information

```{r}
coda_balances_long_predictions_fecal <- data.frame(coda_glmnet_long_fecal$predictions)
colnames(coda_balances_long_predictions_fecal) <- c("Balance")
coda_balances_long_predictions_fecal$treatment <- factor(map_all$treatment)
coda_balances_long_predictions_fecal$week <- factor(map_all$week)
coda_balances_long_predictions_fecal



```

### 8. Create Plots. The first one is a boxplot and scatter plot of microbial balances based on treatment and week. The second one is a bar plot showing genera in the selected balance, filtered based on a coefficient threshold.
# Graph Sample Balances and Species Coefficients (coeff > 0.099)

```{r}
# Species coefficient plot
#plot only the genera that are at least 5% of the contribution to the balance
coda_balance_coeff_10p_fecal <- coda_balance_coeff_fecal[(coda_balance_coeff_fecal$Coefficient >= 0.049 | coda_balance_coeff_fecal$Coefficient<= -0.049),]
coda_balance_coeff_10p_fecal


```
```{r}
# Species coefficient plot
#plot only the genera that are at least 10% of the contribution to the balance
coda_balance_coeff_10p_fecal <- coda_balance_coeff_fecal[(coda_balance_coeff_fecal$Coefficient >= 0.099 | coda_balance_coeff_fecal$Coefficient<= -0.099),]
coda_balance_coeff_10p_fecal


```

```{r}
####
####

p_sample_genera_fecal <- ggplot(
  data = coda_balance_coeff_10p_fecal, 
  aes(x = reorder(Metabolites, Coefficient), y = Coefficient)) +
  geom_bar(stat="identity", width = 0.7, aes(color = Group, fill = Group), show.legend = FALSE) +
  scale_color_manual(values = c("#332288", "#E066A6")) +
  scale_fill_manual(values = c("#332288", "#E066A6")) +
  theme_classic() +
  labs(title = "Fecal Metabolites") +
  xlab(" ") + ylab("Log-contrast Coefficient") +
  theme(text = element_text(size = 15))  +
  theme(plot.title = element_text(size=20, face = "bold",, hjust = 0.5)) +
  coord_flip() 
p_sample_genera_fecal

# Balance prediction plot
p_control_sample_balances_fecal <- ggplot(data = coda_balances_long_predictions_fecal, aes(x = treatment, y = Balance, color = treatment)) +
  geom_line() +
#  geom_point() +
  geom_boxplot() +
  scale_color_manual(values = c("#332288", "#E066A6"))+
  scale_fill_manual(values = c("#332288", "#E066A6")) +
  theme_classic() +
  labs(title = "",
       color = "Group") +  # Change legend title
  xlab(" ") + ylab("Balance Value") +
  theme(text = element_text(size = 15))  +
  theme(plot.title = element_text(size=16)) +
  scale_x_discrete(labels = c("H" = "Half-cup", "T" = "Two-cup")) +
  coord_flip() +
  theme(legend.position = "none")

p_control_sample_balances_fecal

coda_fecal<- egg::ggarrange(p_sample_genera_fecal, p_control_sample_balances_fecal, labels = c("", ""), 
                                          ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(coda_fecal)

```
```{r}

ggsave("fig4_coda_fecal.png", plot = coda_fecal, width = 12, height = 6, units = "in", dpi = 300)




```

Subset rows that only contain two cup samples

```{r}

x_fecal_metabolites_ordered_twocup <- x_fecal_metabolites_ordered[1:76, ]

sub_map <- map_all[1:76, ]

y_twocup <- factor(sub_map$week)


```


## Two cup condition only week 0 vs week 8

```{r}
set.seed(123)

#coda_glmnet_twocup <- coda_glmnet(x_species_ordered ,y, x_time, subject_id, ini_time, end_time, lambda="lambda.min",nfolds=5)

coda_glmnet_fecalmet_twocup <- coda_glmnet(x = x_fecal_metabolites_ordered_twocup, y = y_twocup, lambda = "lambda.min", nvar = NULL, alpha = 0.9, nfolds = 7, showPlots = TRUE, coef_threshold = 0)




```



### 4. Extract Accuracy Measure. The apparent Area Under the Curve (AUC) is extracted as an accuracy measure from the regression results.

```{r}
coda_glmnet_fecalmet_twocup$taxa.num


coda_glmnet_fecalmet_twocup$taxa.name


coda_glmnet_fecalmet_twocup$`log-contrast coefficients`



```





```{r}

coda_glmnet_fecalmet_twocup$`apparent AUC`
coda_glmnet_fecalmet_twocup$`predictions plot`
coda_glmnet_fecalmet_twocup$`signature plot`


```






### 5. Create a Data Frame with Genera and Coefficients. A data frame coda_balance_coeff_control_ocis created to store genera names and their corresponding coefficients from the regression
```{r}

coda_balance_coeff_fecalmet_twocup <- data.frame(cbind(coda_glmnet_fecalmet_twocup$`log-contrast coefficients`, coda_glmnet_fecalmet_twocup$taxa.name))
# view(coda_balance_coeff)

```


### 6. Data Frame Manipulations

```{r}
# Set row names of the data frame coda_balance the taxa names obtained from the regression results
rownames(coda_balance_coeff_fecalmet_twocup) <- coda_balance_coeff_fecalmet_twocup$taxa.name
# Set column names of the data frame to "Coefficient_full" and "Species" for clarity.
colnames(coda_balance_coeff_fecalmet_twocup) <- c("Coefficient_full", "Species")
# Convert "Coefficient_full" column to numeric type, and create a new column "Coefficient" by rounding "Coefficient_full" to two decimal places
coda_balance_coeff_fecalmet_twocup$Coefficient_full <- as.numeric(coda_balance_coeff_fecalmet_twocup$Coefficient_full)
coda_balance_coeff_fecalmet_twocup$Coefficient <-as.numeric(format(round(coda_balance_coeff_fecalmet_twocup$Coefficient_full, 2), nsmall = 2))
# Determine Positive or Negative Coefficients. Create "PosNeg" column based on whether the coefficient is positive or negative; contains the labels "Positive" or "Negative" accordingly.
coda_balance_coeff_fecalmet_twocup$PosNeg <- with(coda_balance_coeff_fecalmet_twocup, ifelse(Coefficient > 0, "Positive", "Negative"))
# Assign Group Labels. The "Group" column is assigned labels "High" or "Low" based on the "PosNeg" column. If the coefficient is positive, it is labeled as "Low"; otherwise, it is labeled as "High".
coda_balance_coeff_fecalmet_twocup$Group <- with(coda_balance_coeff_fecalmet_twocup, ifelse(PosNeg == "Positive", "Low", "High"))
coda_balance_coeff_fecalmet_twocup
```

### 7. Create Data Frame with Samples, Predictions, and Variables. A data frame coda_balances_long_predictions is created to store balance predictions along with sex, group, and sample type information

```{r}
coda_balances_predictions_fecalmet_twocup <- data.frame(coda_glmnet_fecalmet_twocup$predictions)
colnames(coda_balances_predictions_fecalmet_twocup) <- c("Balance")
coda_balances_predictions_fecalmet_twocup$week <- factor(sub_map$week)
coda_balances_predictions_fecalmet_twocup



```

### 8. Create Plots. The first one is a boxplot and scatter plot of microbial balances based on treatment and week. The second one is a bar plot showing genera in the selected balance, filtered based on a coefficient threshold.
# Graph Sample Balances and Species Coefficients (coeff > 0.099)

```{r}
# Species coefficient plot
#plot only the genera that are at least 10% of the contribution to the balance
coda_balance_coeff_10_fecalmet_twocup <- coda_balances_predictions_fecalmet_twocup[(coda_balances_predictions_fecalmet_twocup$Coefficient >= 0.099 | coda_balances_predictions_fecalmet_twocup$Coefficient<= -0.099),]
coda_balance_coeff_10_fecalmet_twocup


```

```{r}

p_sample_genera_twocup_fecalmet <- ggplot(
  data = coda_balance_coeff_10_fecalmet_twocup, 
  aes(x = reorder(Metabolites, Coefficient), y = Coefficient)) +
  geom_bar(stat="identity", width = 0.7, aes(color = Group, fill = Group), show.legend = FALSE) +
  scale_color_manual(values = c("#332288", "#E066A6")) +
  scale_fill_manual(values = c("#332288", "#E066A6")) +
  theme_classic() +
  labs(title = "Fecal Metabolites") +
  xlab(" ") + ylab("Log-contrast Coefficient") +
  theme(text = element_text(size = 15))  +
  theme(plot.title = element_text(size=20, face = "bold",, hjust = 0.5)) +
  coord_flip() 
p_sample_genera_twocup_fecalmet


# Balance prediction plot
p_control_sample_balances_twocup_fecalmet <- ggplot(data = coda_balances_predictions_fecalmet_twocup, aes(x = week, y = Balance, color = week)) +
  geom_line() +
#  geom_point() +
  geom_boxplot() +
  scale_color_manual(values = c("#332288", "#E066A6"))+
  scale_fill_manual(values = c("#332288", "#E066A6")) +
  theme_classic() +
  labs(title = "",
       color = "Group") +  # Change legend title
  xlab(" ") + ylab("Balance Value") +
  theme(text = element_text(size = 15))  +
  theme(plot.title = element_text(size=16)) +
  coord_flip() +
  theme(legend.position = "none")

p_control_sample_balances_twocup_fecalmet

coda_fecal_twocup <- egg::ggarrange(p_sample_genera_twocup_fecalmet, p_control_sample_balances_twocup_fecalmet, labels = c("", ""), 
                                          ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(coda_fecal_twocup)


```



```{r}
####
#### simplify taxa names using simplify_taxa-names_count-tables.R
####

#coda_balance_coeff_control_oc_10p[7,2] <- "Eisenbergiella"

p_sample_genera_twocup_fecalmet <- ggplot(
  data = coda_balance_coeff_10_fecalmet_twocup, 
  aes(x = reorder(Genus, Coefficient), y = Coefficient)) +
  geom_bar(stat="identity", width = 0.7, aes(color = Group, fill = Group), show.legend = FALSE) +
  scale_color_manual(values = c("#332288", "#E066A6")) +
  scale_fill_manual(values = c("#332288", "#E066A6")) +
  theme_classic() +
  labs(title = "Bacteria") +
  xlab(" ") + ylab("Log-contrast Coefficient") +
  theme(text = element_text(size = 15))  +
  theme(plot.title = element_text(size=20, face = "bold",hjust = 0.5)) +
  coord_flip() 
p_sample_genera_twocup_fecalmet

# Balance prediction plot
p_control_sample_balances_twocup_fecalmet <- ggplot(data = coda_balances_predictions_fecalmet_twocup, aes(x = week, y = Balance, color = week)) +
  geom_line() +
  geom_boxplot() +
  scale_color_manual(values = c("#332288", "#E066A6")) +
  scale_fill_manual(values = c("#332288", "#E066A6")) +
  theme_classic() +
  labs(title = "",
       color = "Group") +  # Change legend title
  xlab(" ") + ylab("Balance Value") +
  theme(text = element_text(size = 15))  +
  theme(plot.title = element_text(size=16)) +
  scale_x_discrete(labels = c("0" = "Week 0", "8" = "Week 8")) +
  coord_flip() +
  theme(legend.position = "none")

p_control_sample_balances_twocup_fecalmet

p_coda_bacteria_twocup_fecalmet <- egg::ggarrange(p_sample_genera_twocup_fecalmet, p_control_sample_balances_twocup_fecalmet, labels = c("", ""), 
                                        ncol = 1, nrow = 2, heights = c(1,0.25))
as_ggplot(p_coda_bacteria_twocup_fecalmet)

```
