---
title: "Microbiome-OC"
author: "Fernanda Terrazas"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
Load Packages
```{r Load Packages}
my_packages <- c("FSA", "ggplot2","tidyverse", "corrplot", "ggpubr", "psych","vegan", "nlme", "metap", "caTools", "randomForest", "ROCR", "car", "ape", "tidyr","gridExtra", "lme4") ## "phyloseq"
lapply(my_packages, library, character.only = TRUE)

library("ape")


```


Check for RStudio and packages versions
```{r Load Packages}
# Check RStudio version
library(rstudioapi)
rstudioapi::versionInfo()$version

# Check R packages version
packageVersion("coda4microbiome")
packageVersion("ggplot2")
packageVersion("vegan")
packageVersion("lme4")

```
Set working directory
```{r Load Packages}
setwd("~/microbiome-fruit/data-analyses")

```
SampleID = Unique sample identifier
Treatment = T (two cups of fruit), H (half cup or less of fruit)
Week = 1 (day 1 of cycle), 21 (day 21 of cycle)
Age (years)
Gender = Male or Female
Weight (kg) = from DXA bone density scan (dual-energy x-ray absorption)
Calculated BMI 

# 1. Import data
## 1.1 Import Metadata
```{r}
getwd()
# import file
map_all <- read.csv("metadata-fruit-data-analysis.tsv", header = TRUE, sep = "\t") 

```
## 1.2 ASV table
```{r}
asv_16S_all <- read.delim("~/microbiome-fruit/ASVs/zerofilteredOTU/ASV_table.zerofiltered.tsv", header = TRUE, row.names = 1, check.names = F)

head(asv_16S_all)

#asv_16S <- asv_16S_all[,!(colnames(asv_16S_all) %in% remove_id)]
#head(asv_16S)

```

# Filtering ASVs that are only present in more than 20% of samples

```{r}

# Convert to presence/absence
pres_or_abs <- asv_16S_all > 0

# Prevalence is how many samples each genus is found in
prevalence <- rowSums(pres_or_abs) / ncol(asv_16S_all)

# Filter genera present in more than 20% of samples (gives the names of the ASV)
genera_more_20 <- rownames(asv_16S_all)[prevalence > 0.2]

print(genera_more_20)

# Subset ASV table to keep only prevalent genera
filtered_asv_20 <- asv_16S_all[prevalence > 0.2, ]

head(filtered_asv_20)

```


## 1.3 Import Genus Table
```{r}
genus_all <- read.delim("~/microbiome-fruit/ASVs/zerofilteredOTU/biom_table_genus/genus_table_from_biom.tsv", header = TRUE, row.names = 1, check.names = F,skip=1)


head(genus_all)

```
## 1.3 Import Species Table
```{r}
species_all <- read.delim("~/microbiome-fruit/ASVs/zerofilteredOTU/biom_table_species/species_table_from_biom.tsv", header = TRUE, row.names = 1, check.names = F,skip=1)

head(species_all)

```
## 1.4 Import Alpha Diversity data
```{r}
# alpha_diversity_all <- read.csv("alpha_diversity-zerofiltered.csv", header = TRUE)
shannon <- read.delim("~/microbiome-fruit/diversity/zero-filtered_10k/exported-core-metrics-tables/shannon_vector_alpha-diversity.tsv", header = TRUE)
observed_features <- read.delim("~/microbiome-fruit/diversity/zero-filtered_10k/exported-core-metrics-tables/observed_features_alpha-diversity.tsv", header = TRUE)
faith_pd <- read.delim("~/microbiome-fruit/diversity/zero-filtered_10k/exported-core-metrics-tables/faith_pd_alpha-diversity.tsv", header = TRUE)
pielou_evenness <- read.delim("~/microbiome-fruit/diversity/zero-filtered_10k/exported-core-metrics-tables/evenness_alpha-diversity.tsv", header = TRUE)

# Rename the sample ID column in each data frame so it matches on all data frames
colnames(shannon)[1] <- "SampleID"
colnames(observed_features)[1] <- "SampleID"
colnames(faith_pd)[1] <- "SampleID"
colnames(pielou_evenness)[1] <- "SampleID"

# Merge the data frames
alpha_diversity_all <- merge(shannon, observed_features, by = "SampleID")
alpha_diversity_all <- merge(alpha_diversity_all, faith_pd, by = "SampleID")
alpha_diversity_all <- merge(alpha_diversity_all, pielou_evenness, by = "SampleID")

# only for consistency:
alpha_diversity <- alpha_diversity_all

```
## 1.5 Import Unifrac Matrices
exported from qiime2
```{r}
weighted_unfiltered <- read.csv("~/microbiome-fruit/diversity/zero-filtered_10k/exported-core-metrics-tables/weighted_unifrac_distance-matrix.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = F)
head(weighted_unfiltered)

```

```{r}
unweighted_unfiltered <- read.csv("~/microbiome-fruit/diversity/zero-filtered_10k/exported-core-metrics-tables/unweighted_unifrac_distance-matrix.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = F)
head(unweighted_unfiltered)

```
check dimensions
```{r}
dim(weighted_unfiltered)
dim(unweighted_unfiltered)

```
## 1.6 Import taxonomy table into R and format to work with phyloseq
```{r}
# Read the taxonomy table (assuming it's in a CSV or TSV format):
taxonomy_raw <- read.table("~/microbiome-fruit/taxonomy_export/taxonomy.tsv", stringsAsFactors = FALSE, header = TRUE, sep = "\t")
head(taxonomy_raw)

# Remove prefixes from Taxon names
taxonomy_raw$Taxon <- gsub("d__|p__|c__|o__|f__|g__|s__", "", taxonomy_raw$Taxon)
head(taxonomy_raw)

```
Split the taxonomic ranks and create a new taxonomy table:
```{r}
# # Convert the 'Taxon' column to a character before splitting (may not need to do this)
# taxonomy_raw$Taxon <- as.character(taxonomy_raw$Taxon)

# Split the "Taxon" column into taxonomic ranks
tax_ranks <- strsplit(taxonomy_raw$Taxon, "; ")

# Create a data frame with separate columns for each taxonomic rank
tax_table <- do.call(rbind, lapply(tax_ranks, function(x) {
  c(Domain = x[1], Phylum = x[2], Class = x[3], Order = x[4], Family = x[5], Genus = x[6], Species = x[7])
}))

# Set the row names of the tax_table to match the ASV/OTU names
rownames(tax_table) <- taxonomy_raw$Feature.ID
head(tax_table)

```
## 1.7 Import phylogenetic tree into R
```{r}
# Use read.tree from the 'ape' package to import Newick files
phylo_tree_rooted <- read.tree("~/microbiome-fruit/phylogeny/exported-rooted-tree/rooted-tree.nwk")
phylo_tree_unrooted <- read.tree("~/microbiome-fruit/phylogeny/exported-unrooted-tree/unrooted-tree.nwk")

```
## 1.8 Subset metadata into Treatment and Cycle day groups
```{r}
# Check number of samples per group
table(map_all$treatment, map_all$week)

# Create each group to use for alpha diversity and subsequent analyses

# Subset the columns - treatment and week (time point)
group  <- subset(map_all, treatment %in% c("T", "H") & week %in% c("0","8"))
group <- map_all[, c("treatment", "week")]
table(group$treatment, group$week)
view(group)

```
